Retirement Consumption: Evidence from UK Pension Reform

### 2023-07-27 Texfiles/Draft1.tex
last before name change

        Word count: 8335\footnote{This is an estimate that my LaTeX editor gave me.} + 1750 (figure and table word equivalent)
        \vfill
        I confirm that this is entirely my own work and has not previously been submitted for assessment, and I have read and understood the University’s and Faculty’s definition of Plagiarism
    \end{center}
\end{titlepage}
\newpage
\maketitle
\begin{abstract}
    I identify the consumption responses of retirees to the UK's `pensions
    freedom act', a landmark law change that meant defined contribution pension
    holders were no longer forced to annuitise their funds and could instead
    access their pensions in a variety of ways. I use a discontinuity in year of
    retirement and a difference-in-difference approach, harnessing those who
    were unaffected by the policy reform as a natural control group. I then
    solve three sets of lifecycle models, tracking assets from retirement until an
    agent reaches age 110. I simulate their consumption with and without
    annuities, mimicking the policy reform, and match their consumption
    responses to what was seen in the empirical models. I find that a lifecycle
    model with a bequest motive best fits the identified consumption response —
    a result that could inform policymakers in designing optimal pension
    policies for an ageing population.
\end{abstract}
\newpage
\tableofcontents
\newpage

@@

Columns 2 through 4 use food consumption as the outcome variable. Interestingly,
food consumption inside the house does not increase as a result of the policy
reform but food consumed away from the house does increase. A shock to income,
caused by no longer needing to annuitise, may cause spending on luxury goods to
increase and for spending on their substitutes, like food in the house relative
to food away from home, to decrease. This wealth effect may be driving the
results we observe in Table \ref{tab:DcOnlyRes} since food consumption outside the
home increases as does expenditure on clothing -- seen in column 5.

@@

The key identifying assumption for DD is the parallel trends assumption. In this
context, it means that consumption differences between the DB, DC and no pension
group across cohorts would have been the same were it not for the policy change.
Therefore for DD to be invalid, we would need a reason to expect that the
individuals with a DB and no pension who are retiring post 2014 are impacted by
some factor that did not equally impact the DC group. Given that there were no
other pension reforms at this time that differently affected these groups, I
claim that the parallel trends assumption holds.

@@

Using the policy function at age 110, which we just found by doing the
optimisation at age 110, I calculate the value for all income and asset states
at age 110. This value is then used to solve the problem an individual at age
109 faces, specifically, for each income and asset state, they need to find the
assets next period that maximise: utility from consumption; bequests multiplied
by the probability of death; and thevalue function at age 110, discounted by
$\beta$, multiplied by the probability of being alive. I repeat this process
back to the age of retirement to obtain optimal consumption amounts for each
year of retirement and the associated value functions. I write code in Julia
that solves this maximisation problem.

@@

Where $C$ is the one-off payment, $\delta$ is the `money's worth' of an annuity
and $p_{t|Retage}$ is the probability of death at age $t$ conditional on being
age $Retage$. So individuals can move down $C$ on the asset grid for gaining
$Ann$ on the permanent income grid for the rest of their lives. The `money's
worth' of an annuity is the ratio of expected present discounted value to the
price of an annuity that an individual can purchase. An actuarially fair annuity
will have a money's worth of 1 whilst an annuity that pays more than it costs
will have a money's worth greater than 1. Generally they have been found to be
between $0.8$ and $0.9$ for the population as a whole in the UK
\citep{finkelstein_porteba_2002, finkelstein_porteba_2004, mitchell_et_al_1999}.

@@

Future research is still needed to understand the full impact the reform has had
on consumption across the whole of retirement, especially as more data on
retirees impacted by the reform becomes available. The effect that reforms since
this one, such as allowing individuals on DB plans to switch to DC plans, has
had on consumption could also be studied. Moreover, as the number of individuals
retiring with a DC pension grows, understanding the mechanisms
behind their choices will matter more to policy makers. For example, if the
annuitisation puzzle is caused by a bequest motive rather than low money's worth
or short subjective life expectancies, then policies that force annuitisation or
tax assets to fund state pensions will not be welfare enhancing.

>>>>

        Word count: 8336\footnote{This is an estimate that my LaTeX editor gave me.} + 1650 (figure and table word equivalent)
        \vfill
        I confirm that this is entirely my own work and has not previously been submitted for assessment, and I have read and understood the University’s and Faculty’s definition of Plagiarism
    \end{center}
\end{titlepage}
\newpage
\maketitle
\begin{abstract}
    I identify the consumption responses of retirees to the UK's `pensions
    freedom act', a landmark law change that meant defined contribution pension
    holders were no longer forced to annuitise their funds and could instead
    access their pensions in a variety of ways. I use a discontinuity in year of
    retirement and a difference-in-difference approach, harnessing those who
    were unaffected by the policy reform as a natural control group. I then
    solve three sets of lifecycle models, tracking assets from retirement until an
    agent reaches age 110. I simulate their consumption with and without
    annuities, mimicking the policy reform, and match their consumption
    responses to what was seen in the empirical models. I find that a lifecycle
    model with a bequest motive best fits the identified consumption response —
    a result that could inform policymakers in designing optimal pension
    policies for an ageing population.
\end{abstract}
\newpage
\tableofcontents
\newpage

@@

Columns 2 through 4 use food consumption as the outcome variable. Interestingly,
food consumption inside the house does not increase as a result of the policy
reform but food consumed away from the house does increase. A shock to income,
caused by no longer needing to annuitise, may cause spending on luxury goods to
increase and for spending on their substitutes, like food in the house relative
to food away from home, to decrease. This wealth effect may be driving the
results we observe in Table \ref{tab:DcOnlyRes}, since food consumption outside the
home increases as does expenditure on clothing -- seen in column 5.

@@

The key identifying assumption for DD is the parallel trends assumption. In this
context, it means that consumption differences between the DB, DC and no pension
groups across cohorts would have been the same were it not for the policy change.
Therefore for DD to be invalid, we would need a reason to expect that the
individuals with a DB and no pension who are retiring post 2014 are impacted by
some factor that did not equally impact the DC group. Given that there were no
other pension reforms at this time that differently affected these groups, I
claim that the parallel trends assumption holds.

@@

Using the policy function at age 110, which we just found by doing the
optimisation at age 110, I calculate the value for all income and asset states
at age 110. This value is then used to solve the problem an individual at age
109 faces. Specifically, for each income and asset state, they need to find the
assets next period that maximise: utility from consumption; bequests multiplied
by the probability of death; and the value function at age 110, discounted by
$\beta$, multiplied by the probability of being alive. I repeat this process
back to the age of retirement to obtain optimal consumption amounts for each
year of retirement and the associated value functions. I write code in Julia
that solves this maximisation problem.

@@

Where $C$ is the one-off payment, $\delta$ is the `money's worth' of an annuity
and $p_{t|Retage}$ is the probability of death at age $t$ conditional on being
age $Retage$. So individuals can move down $C$ on the asset grid for gaining
$Ann$ on the permanent income grid for the rest of their lives. The `money's
worth' of an annuity is the ratio of expected present discounted value to the
price of an annuity that an individual can purchase. An actuarially fair annuity
will have a money's worth of 1 whilst an annuity that pays more than it costs
will have a money's worth greater than 1. Generally, they have been found to be
between $0.8$ and $0.9$ for the population as a whole in the UK
\citep{finkelstein_porteba_2002, finkelstein_porteba_2004, mitchell_et_al_1999}.

@@

Future research is still needed to understand the full impact the reform has had
on consumption across the whole of retirement, especially as more data on
retirees impacted by the reform becomes available. The effect that reforms since
this one, such as allowing individuals on DB plans to switch to DC plans, has
had on consumption could also be studied. Moreover, as the number of individuals
retiring with a DC pension grows, understanding the mechanisms
behind their choices will matter more to policy makers. For example, if the
annuitisation puzzle is caused by a bequest motive rather than low money's worth
or short subjective life expectancies, then policies that force annuitisation or
tax assets to fund state pensions will not be welfare-enhancing.
